Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-06-21 paper/sec/invariants.tex
Explain where inequality i:subsample_condition comes from.
Note: This is due to reviewer A comment:
> l261: What is "f"? I generally have some trouble understanding where (5) comes from

Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$. By~\cref{eq:integral_bind},
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \integral{\Omega}{\integral{\Ber(p)}{p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right)}{\tau}}{\sigma} \textrm{.}
    % \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
\]
Note that we write $p$ or $\chi$ even though we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
To see that the right-hand side is equal to $1$, it is useful to consider cases on whether $a=s$.
When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.

The same invariant-based argument is possible for $\mathrm{step}_2$.
Let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
Then by~\cref{eq:integral_bind},
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
    \integral{\Omega}
      {\left(\ift{|\chi|=n}{\left(\integral{\mathrm{subsample}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}\right)}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}
    \textrm{.}
\]
Note that the \emph{true} and \emph{false} cases of the inner \textbf{if-then-else} both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} \indicat(s \in \chi)$.
Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
This completes the invariance proof for~\cref{eq:indicator_eq}.

>>>>

Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$. By~\cref{eq:integral_bind},
\begin{equation}\label{eq:step_1_exp}
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \integral{\Omega}{\integral{\Ber(p)}{p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right)}{\tau}}{\sigma} \textrm{.}
    % \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
\end{equation}
Note that we write $p$ or $\chi$ even though we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
To see that the right-hand side is equal to $1$, it is useful to consider cases on whether $a=s$.
When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.

The same invariant-based argument is possible for $\mathrm{step}_2$.
Let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
Then by~\cref{eq:integral_bind}, $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1}X_s]$ equals
\begin{equation}\label{eq:step_2_exp}
%  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right]
    \integral{\Omega}
      {\left(\ift{|\chi|=n}{\left(\integral{\mathrm{subsample}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}\right)}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}
    \textrm{.}
\end{equation}
Note that the \emph{true} and \emph{false} cases of the inner \textbf{if-then-else} both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} \indicat(s \in \chi)$.
Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
This completes the invariance proof for~\cref{eq:indicator_eq}.

### 2025-06-21 paper/sec/negdep.tex
Explain where inequality i:subsample_condition comes from.
Note: This is due to reviewer A comment:
> l261: What is "f"? I generally have some trouble understanding where (5) comes from

When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation:
\begin{equation}\label[ineq]{i:subsample_condition}
  \integral{\mathrm{subsample}(\chi)}{\prod_{s \in S} g(\indicat(s \in \tau))}{\tau} \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
\end{equation}
for all non-negative functions $g$ and any $S \subseteq \chi$.
Any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.

>>>>

When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation.
It should be noted that the condition becomes apparent while establishing \cref{i:func_invariant} using similar but more general steps as described in Equations~\ref{eq:step_1_exp} and \ref{eq:step_2_exp}.%
\footnote{A step-by-step proof of the derivation is also available in the appendix of the formalization~\cite[Appendix~A]{CVM_Distinct_Elements-AFP}.}
\begin{equation}\label[ineq]{i:subsample_condition}
  \integral{\mathrm{subsample}(\chi)}{\prod_{s \in S} g(\indicat(s \in \tau))}{\tau} \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
\end{equation}
for all non-negative functions $g$ and any $S \subseteq \chi$.
Any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
